---
title: Writing efficient R code
layout: default
---

```{r, include=FALSE}
library(rbenchmark)
library(microbenchmark)
```

## 1. Pre-allocate memory

It is very inefficient to iteratively add elements to a vector, matrix,
data frame, array or list (e.g., using *c*, *cbind*,
*rbind*, etc. to add elements one at a time) in R. (Note that Python handles this sort of thing much better.) Instead, create the full object in advance
(this is equivalent to variable initialization in compiled languages)
and then fill in the appropriate elements. The reason is that when
R appends to an existing object, it creates a new copy and as the
object gets big, most of the computation involves the repeated 
memory allocation to create the new objects.  Here's
an illustrative example, but of course we would not fill a vector
like this using loops because we would in practice use vectorized calculations.

```{r, preallocate, cache=TRUE}
library(rbenchmark)

n <- 10000
z <- rnorm(n)

fun_append <- function(vals) {
   x <- exp(vals[1])
   n <- length(vals) 
   for(i in 2:n) x <- c(x, exp(vals[i]))
   return(x)
}
fun_prealloc <- function(vals) {
   n <- length(vals)
   x <- rep(as.numeric(NA), n)
   for(i in 1:n) x[i] <- exp(vals[i])
   return(x)
}
fun_vec <- function(vals) {
  x <- exp(vals)
  return(x)
}
benchmark(fun_append(z), fun_prealloc(z), fun_vec(z),
replications = 20, columns=c('test', 'elapsed', 'replications'))
```

It's not necessary to use *as.numeric* above though it saves
a bit of time. **Challenge**: figure out why I have `as.numeric(NA)`
and not just `NA`. Hint: what is the type of `NA`?

In some cases, we can speed up the initialization by initializing a vector of length one and then changing its length and/or dimension, although in  many practical
circumstances this would be overkill.

For example, for matrices, start with a vector of length one, change the length, and then change the
dimensions

```{r, init-matrix}
nr <- nc <- 2000
benchmark(
   x <- matrix(as.numeric(NA), nr, nc),
   {x <- as.numeric(NA); length(x) <- nr * nc; dim(x) <- c(nr, nc)},
replications = 10, columns=c('test', 'elapsed', 'replications'))
```

For lists, we can do this

```{r, init-list}
myList <- vector("list", length = n)
```



## 2. Vectorized calculations

One key way to write efficient R code is to take advantage of R's
vectorized operations.

```{r, vectorize, cache=TRUE}
n <- 1e6
x <- rnorm(n)
benchmark(
    x2 <- x^2,
    { x2 <- as.numeric(NA)
      length(x2) <- n
      for(i in 1:n) { x2[i] <- x[i]^2 } },
    replications = 10, columns=c('test', 'elapsed', 'replications'))
```

So what is different in how R handles the calculations above that
explains the huge disparity in efficiency? The vectorized calculation is being done natively
in C in a for loop. The explicit R for loop involves executing the for
loop in R with repeated calls to C code at each iteration. This involves a lot
of overhead because of the repeated processing of the R code inside the loop. For example,
in each iteration of the loop, R is checking the types of the variables because it's possible
that the types might change, such as in this loop:

```
x <- 3
for( i in 1:n ) {
     if(i == 7) {
          x <- 'foo'
     }
     y <- x^2
}
```

You can
usually get a sense for how quickly an R call will pass things along
to C or Fortran by looking at the body of the relevant function(s) being called
and looking for *.Primitive*, *.Internal*, *.C*, *.Call*,
or *.Fortran*. Let's take a look at the code for `+`,
*mean.default*, and *chol.default*. 

```{r, primitive}
`+`
mean.default
chol.default
```

Many R functions allow you to pass in vectors, and operate on those
vectors in vectorized fashion. So before writing a for loop, look
at the help information on the relevant function(s) to see if they
operate in a vectorized fashion. Functions might take vectors for one or more of their arguments.
Here we see that `nchar` is vectorized and that various arguments to `substring` can be vectors.

```{r, vectorized}
address <- c("Four score and seven years ago our fathers brought forth",
             " on this continent, a new nation, conceived in Liberty, ",
             "and dedicated to the proposition that all men are created equal.")
nchar(address)
# use a vector in the 2nd and 3rd arguments, but not the first
startIndices = seq(1, by = 3, length = nchar(address[1])/3)
startIndices
substring(address[1], startIndices, startIndices + 1)
```

**Challenge**: Consider the chi-squared statistic involved in
a test of independence in a contingency table:

\[
\chi^{2}=\sum_{i}\sum_{j} (y_{i,j}-e_{i,j})^{2} / e_{i,j},\,\,\,\, e_{i,j}= ( y_{i\cdot}y_{\cdot j} ) / y_{\cdot\cdot}
\]

where $y_{i\cdot}=\sum_{j}y_{i,j}$ and $y_{\cdot j} = \sum_{i} y_{i,j}$ and $y_{\cdot\cdot} = \sum_{i} \sum_{j} y_{i,j}$. Write this in a vectorized way
without any loops.  Note that 'vectorized' calculations also work
with matrices and arrays.

Vectorized operations can sometimes be faster than built-in functions 
(note here the *ifelse* is notoriously slow),
and clever vectorized calculations even better, though sometimes the
code is uglier. Here's an example of setting all negative values in a 
vector to zero.

```{r, vec-tricks, cache=TRUE}
x <- rnorm(1000000)
benchmark(
   truncx <- ifelse(x > 0, x, 0),
   {truncx <- x; truncx[x < 0] <- 0},
   truncx <- x * (x > 0),
   replications = 10, columns=c('test', 'elapsed', 'replications'))
```


Additional tips:

  - If you do need to loop over dimensions of a matrix or array, if possible
loop over the smallest dimension and use the vectorized calculation
on the larger dimension(s). For example if you have a 10000 by 10 matrix, try to set
up your problem so you can loop over the 10 columns rather than the 10000 rows.
  - In general, looping over columns is likely to be faster than looping over rows
given R's column-major ordering (matrices are stored in memory as a long array in which values in a column are adjacent to each other) (see more in Section 4.6 on the cache).
  - You can use direct arithmetic operations to add/subtract/multiply/divide
a vector by each column of a matrix, e.g. `A*b` does element-wise multiplication of
each column of *A* by a vector *b*. If you need to operate
by row, you can do it by transposing the matrix. 

Caution: relying on R's recycling rule in the context of vectorized
operations, such as is done when direct-multiplying a matrix by a
vector to scale the rows relative to each other, can be dangerous as the code is not transparent
and poses greater dangers of bugs. In some cases you may want to
first write the code transparently and
then compare the more efficient code to make sure the results are the same. It's also a good idea to  comment your code in such cases.

## 3. Using *apply* and specialized functions

Historically, another core efficiency strategy in R has been to use the *apply* functionality (e.g., `apply`, `sapply`, `lapply`, `mapply`, etc.).

### Some faster alternatives to `apply`

Note that even better than *apply* for calculating sums or means of columns
or rows (it also can be used for arrays) is `rowSums`,`colSums`,`rowMeans`, and `colMeans`.

```{r, apply, cache=TRUE}
n <- 3000; x <- matrix(rnorm(n * n), nr = n)
benchmark(
   out <- apply(x, 1, mean),
   out <- rowMeans(x),
   replications = 10, columns=c('test', 'elapsed', 'replications'))
```

We can 'sweep' out a summary statistic, such as subtracting
off a mean from each column, using *sweep*

```{r, sweep}
system.time(out <- sweep(x, 2, STATS = colMeans(x), FUN = "-"))
```

Here's a trick for doing the sweep based on vectorized calculations, remembering
that if we subtract a vector from a matrix, it subtracts each element
of the vector from all the elements in the corresponding ROW. Hence the 
need to transpose twice. 

```{r, vectorized-sweep}
system.time(out2 <- t(t(x) - colMeans(x)))
identical(out, out2)
```

### Are *apply*, *lapply*, *sapply*, etc. faster than loops?

Using *apply* with matrices and versions of *apply* with lists or vectors (e.g., `lapply`, `sapply`) may or may not be faster
than looping but generally produces cleaner code.

Whether looping and use of apply variants is slow will depend in part on whether a substantial part of the work is
in the overhead involved in the looping or in the time required by the function
evaluation on each of the elements. If you're worried about speed,
it's a good idea to benchmark the *apply* variant against looping.

Here's an example where *apply* is not faster than a loop. Similar
examples can be constructed where *lapply* or *sapply* are not faster
than writing a loop. 

```{r, apply-vs-for}
n <- 500000; nr <- 10000; nCalcs <- n/nr
mat <- matrix(rnorm(n), nrow = nr)
times <- 1:nr
system.time(
  out1 <- apply(mat, 2, function(vec) {
                         mod = lm(vec ~ times)
                         return(mod$coef[2])
                     })) 
system.time({
                out2 <- rep(NA, nCalcs)
                for(i in 1:nCalcs){
                    out2[i] = lm(mat[ , i] ~ times)$coef[2]
                }
            }) 
```

And here's an example, where (unlike the previous example) the core computation is very fast, so we might expect the overhead of looping to be important. I believe that in old versions of R the *sapply* in this  example was faster than looping in R, but that doesn't seem to be the case currently.
I think this may be related to various somewhat recent improvements in R's handling of loops, possibly including the use of the byte compiler.

```{r, apply-vs-for-part2, cache=TRUE}
z <- rnorm(1e6)
fun_loop <- function(vals) {
    x <- as.numeric(NA)
    n <- length(vals)
    length(x) <- n
    for(i in 1:n) x[i] <- exp(vals[i])
    return(x)
}

fun_sapply <- function(vals) {
    x <- sapply(vals, exp)
    return(x)
}

fun_vec <- function(vals) {
    x <- exp(vals)
    return(x)
}

benchmark(fun_loop(z), fun_sapply(z), fun_vec(z),
replications = 10, columns=c('test', 'elapsed', 'replications'))
```

You'll notice if you look at the R code for *lapply* (*sapply* just calls *lapply*) that it calls directly out to C code, so the for loop is executed in compiled code. However, the code being executed at each iteration is still R code, so there is still all the overhead of the R interpreter. 

```{r, lapply-callout}
print(lapply)
```

## 4. Matrix algebra efficiency

Often calculations that are not explicitly linear algebra calculations
can be done as matrix algebra. If our R installation has a fast (and possibly parallelized) BLAS, this allows our calculation to take advantage of it.

For example, we can sum the rows of a matrix by multiplying by a vector of ones. Given the extra computation involved in actually multiplying each number by one, it's surprising that this is faster than using R's heavily optimized *rowSums* function. 

```{r, matrix-calc}
mat <- matrix(rnorm(500*500), 500)
benchmark(apply(mat, 1, sum),
	mat %*% rep(1, ncol(mat)),
	rowSums(mat),
	replications = 10, columns=c('test', 'elapsed', 'replications'))
```

On the other hand, big matrix operations can be slow.

> **Challenge**: Suppose you want a new matrix that computes the differences between successive columns of a matrix of arbitrary size. How would you do this as matrix algebra operations? It's possible to write it as multiplying the matrix by another matrix that contains 0s, 1s, and -1s in appropriate places.  Here it turns out that the *for* loop is much faster than matrix multiplication. However, there is a way to do it faster as matrix direct subtraction. 

### Order of operations and efficiency

When doing matrix algebra, the order in which you do operations can
be critical for efficiency. How should I order the following calculation?

```{r, linalg-order, cache=TRUE}
n <- 5000
A <- matrix(rnorm(5000 * 5000), 5000)
B <- matrix(rnorm(5000 * 5000), 5000)
x <- rnorm(5000)
system.time(
  res1 <- A %*% B %*% x
)
system.time(
  res2 <- A %*% (B %*% x)
)
```

Why is the second order much faster?

### Avoiding unnecessary operations 

We can use the matrix direct product (i.e., `A*B`) to do
some manipulations much more quickly than using matrix multiplication.
**Challenge**: How can I use the direct product to find the trace
of a matrix, `trace(XY)`? 

Finally, when working with diagonal matrices, you can generally get much faster results by being smart. The following operations: `X+D`, `DX`, `XD`
are mathematically the sum of two matrices and products of two matrices.
But we can do the computation without using two full matrices.
**Challenge**: How?

```{r, diag, cache=TRUE}
n <- 1000
X <- matrix(rnorm(n^2), n) 
diagvals <- rnorm(n)
D = diag(diagvals)
# the following lines are very inefficient
summedMat <- X + D
prodMat1 <- D %*% X
prodMat2 <- X %*% D
# How can we do each of those operations much more quickly?
```

More generally, sparse matrices and structured matrices (such as block
diagonal matrices) can generally be worked with MUCH more efficiently
than treating them as arbitrary matrices. The R packages *spam* (for arbitrary
sparse matrices), *bdsmatrix* (for block-diagonal matrices),
and *Matrix* (for a variety of sparse matrix types) can help, as can specialized code available in other languages,
such as C and Fortran packages.


## 5. Fast mapping/lookup tables

Sometimes you need to map between two vectors. E.g., 
$y_{i}\sim\mathcal{N}(\mu_{j[i]},\sigma^{2})$
is a basic ANOVA type structure, where multiple observations
are associated with a common mean, $\mu_j$, via the `j[i]` mapping. 

How can we quickly look up the mean associated with each observation?
A good strategy is to create a vector, `grp`, that gives a numeric
mapping of the observations to their cluster, playing the role of `j[i]` above. Then you can access
the $\mu$ value relevant for each observation as: `mus[grp]`. This requires
that `grp` correctly map to the right elements of `mus`.

The *match* function can help in creating numeric indices that can then be used for lookups. 
Here's how you would create an index vector, *grp*, if it doesn't already exist.

```{r, match-lookup}
df <- data.frame(
    id = 1:5,
    clusterLabel = c('C', 'B', 'B', 'A', 'C')) 
info <- data.frame(
    grade = c('A', 'B', 'C'),
    numGrade = c(95, 85, 75),
    fail = c(FALSE, FALSE, TRUE) )
grp <- match(df$clusterLabel, info$grade) 
df$numGrade <- info$numGrade[grp]
df
```

### Lookup by name versus index

In the example above we looked up the `mu` values based on `grp`, which supplies the needed indexes as numeric indexes.

R also allows you to look up elements of vector by name, as illustrated here by rearranging the code above a bit:

```{r, name-lookup}
info2 <- info$numGrade
names(info2) <- info$grade
info2
info2[df$clusterLabel]
```

You can do similar things in terms of looking up by name with dimension
names of matrices/arrays, row and column names of dataframes, and
named lists.

However, looking things up by name can be slow relative to looking up by index.
Here's a toy example where we have a vector or list with 1000 elements and
the character names of the elements are just the character versions of the 
indices of the elements.  

```{r, index-lookup}
library(microbenchmark)

n <- 1000
x <- 1:n
xL <- as.list(x)
nms <- paste0("var", as.character(x))
names(x) <- nms
names(xL) <- nms
x[1:3]
xL[1:3]
microbenchmark(
    x[500],  # index lookup in vector
    x["var500"], # name lookup in vector
    xL[[500]], # index lookup in list
    xL[["var500"]]) # name lookup in list
```

Lookup by name is slow because R needs to scan through the objects
one by one until it finds the one with the name it is looking for.
In contrast, to look up by index, R can just go directly to the position of interest.

Side note: there is a lot of variability across the 100 replications shown above. This might have to do with cache effects (see next section).

In contrast, we can look up by name in an environment very quickly, because environments in R use hashing, which allows for fast lookup that does not require scanning through all of the names in the environment. In fact, this is how R itself looks for values when you specify variables in R code, because the global environment, function frames, and package namespaces are all environments. 

```{r, env-lookup, cache=TRUE}
xEnv <- as.environment(xL)  # convert from a named list
xEnv$var500  
microbenchmark(
  x[500],
  xL[[500]],
  xEnv[["var500"]],
  xEnv$var500
)
```


## 6. Cache-aware programming

In addition to main memory (what we usually mean when we talk about RAM), computers also have memory caches, which are small amounts of fast memory that can be accessed very quickly by the processor. For example your computer might have L1, L2, and L3 caches, with L1 the smallest and fastest and L3 the largest and slowest. The idea is to try to have the data that is most used by the processor in the cache. 

If the next piece of data needed for computation is available in the cache, this is a *cache hit* and the data can be accessed very quickly. However, if the data is not available in the cache, this is a *cache miss* and the speed of access will be a lot slower. *Cache-aware programming* involves writing your code to minimize cache misses. Generally when data is read from memory it will be read in chunks, so values that are contiguous will be read together.

How does this inform one's programming? For example, if you have a matrix of values stored in column-major order, computing on a column will be a lot faster than computing on a row, because the column can be read into the cache from main memory and then accessed in the cache. In contrast, if the matrix is large and therefore won't fit in the cache, when you access the values of a row, you'll have to go to main memory repeatedly to get the values for the row because the values are not stored contiguously.

There's a nice example of the importance of the cache at [the bottom of this blog post](https://wrathematics.github.io/2016/10/28/comparing-symmetric-eigenvalue-performance/).

If you know the size of the cache, you can try to design your code so that in a given part of your code you access data structures that will fit in the cache. This sort of thing is generally more relevant if you're coding in a language like C. But it can matter sometimes in R too. Here's an example:

```{r, cache-aware, cache=TRUE}
nr <- 800000
nc <- 100
## large matrix that won't fit in cache
A <- matrix(rnorm(nr * nc), nrow = nr)
tA <- t(A)
benchmark(
    apply(A, 2, mean),   ## operate by column
    apply(tA, 1, mean),  ## exact same calculation, but by row
    replications = 10, columns=c('test', 'elapsed', 'replications'))
```

Now let's compare things when we make the matrix small enough that it fits in the cache. In this case it fits into the largest (L3) cache but not the smaller (L2 and L1) caches, and we see that the difference in speed disappears.

```{r, cache-aware2, cache=TRUE}
nr <- 800
nc <- 100
## small matrix that should fit in cache
A <- matrix(rnorm(nr * nc), nrow = nr)
## Yep, the size is less than the L3 cache:
object.size(A)
memuse::Sys.cachesize()

tA <- t(A)
benchmark(apply(A, 2, mean),  ## by column
          apply(tA, 1, mean),  ## by row
  replications = 10, columns=c('test', 'elapsed', 'replications'))

```

