Writing Efficient R Code
==============================================================

Chris Paciorek, Department of Statistics, UC Berkeley

```{r setup, include=FALSE}
library(knitr)
library(stringr)
read_chunk('efficient-R.R')
```

# 0) This Tutorial

This tutorial covers strategies for writing efficient R code by taking advantage of the underlying structure of how R works. In addition it covers tools and strategies  for timing and profiling R code.

In the future, a screencast of the material may accompany this document.

You should be able to work through this tutorial in any working R installation, including through RStudio. To work through it using R linked to a fast linear algebra package, you may want to use a virtual machine developed here at Berkeley, the [Berkeley Common Environment (BCE)](http://bce.berkeley.edu). BCE is a virtual Linux machine - basically it is a Linux computer that you can run within your own computer, regardless of whether you are using Windows, Mac, or Linux. This provides a common environment so that things behave the same for all of us. 

This tutorial assumes you have a working knowledge of R. 

Materials for this tutorial, including the R markdown file and associated code files that were used to create this document are available on Github at (https://github.com/berkeley-scf/tutorial-efficient-R).  You can download the files by doing a git clone from a terminal window on a UNIX-like machine, as follows:
```{r, clone, eval=FALSE}
git clone https://github.com/berkeley-scf/tutorial-efficient-R
```

To create this HTML document, simply compile the corresponding R Markdown file in R as follows (the following will work from within BCE after cloning the repository as above).
```{r, build-html, eval=FALSE}
Rscript -e "library(knitr); knit2html('efficient-R.Rmd')"
```
This tutorial by Christopher Paciorek is licensed under a Creative Commons Attribution 3.0 Unported License.


# 1) Background


In part because R is an interpreted language and in part because R
is very dynamic (objects can be modified essentially arbitrarily after
being created), R can be slow. Hadley Wickham's Advanced R book has
a section called *Performance* that discusses this in detail. However, there
are a variety of ways that one can write efficient R code.

In general, try to make use of R's built-in functions (including matrix
operations and linear algebra), as these tend to be implemented
internally (i.e., via compiled code in C or Fortran).  Sometimes you
can figure out a trick to take your problem and transform it to make
use of the built-in functions.

Before you spend a lot of time trying to make your code go faster,
it's best to first write transparent, easy-to-read code to help avoid
bugs.  Then if it doesn't run fast enough, time the different parts of
the code (profiling) to assess where the bottlenecks are. Concentrate
your efforts on those parts of the code. Try out different
specifications, checking that the results are the same as your
original code.  And as you gain more experience, you'll get some
intuition for what approaches might improve speed, but even with
experience I find myself often surprised by what matters and what
doesn't.

Section 2 of this document discusses the use of fast linear algebra libraries, Section 3 discusses tools for timing and profiling code, and Section 4 discusses core strategies for writing efficient R code.

# 2) Fast linear algebra

One way to speed up a variety of operations in R (sometimes by as much as an order of magnitude) is to make sure your installation of R uses an optimized BLAS (Basic Linear Algebra Subroutines). The BLAS underlies all linear algebra, including costly calculations such as matrix-matrix multiplication and matrix decompositions such as the SVD and Cholesky decomposition. Some optimized BLAS packages are:
 - Intel's *MKL*
 - *OpenBLAS*
 - AMD's *ACML*
 - *vecLib* for Macs

To use an optimized BLAS, talk to your systems adminstrator, see [Section A.3 of the R Installation and Administration Manual](https://cran.r-project.org/manuals.html), or see [these instructions to use *vecLib* BLAS on your own Mac](http://statistics.berkeley.edu/computing/blas).

Any calls to BLAS or to the LAPACK libraries that use BLAS to do higher-level linear algebra calculations will be nearly as fast as if you used C/C++ or Matlab, because R is using the compiled code from the BLAS and LAPACK libraries. 

In addition, the BLAS libraries above are threaded -- they can use more than one core, and often will do so by default. More details in the tutorial on parallel programming. 

 
# 3) Tools for assessing efficiency


## 3.1) Benchmarking

*system.time* is very handy for comparing the speed of different
implementations. Here's a basic comparison of the time to calculate the row means of a matrix using a for loop compared to the built-in *rowMeans* function.

```{r, system-time}
```



To time code that runs very quickly, you may want to use the *microbenchmark*
package. Of course one would generally only care about such timing if a larger operation does the quick calculation very many times. Here's a comparison of different ways of accessing an element of a dataframe.

```{r, microbenchmark}
```


The *rbenchmark* package provides a nice wrapper function, *benchmark*,
that automates timings and comparisons. 

```{r, benchmark}
```

In general, it's a good idea to repeat (replicate) your timing, as there is some stochasticity in how fast your computer will run a piece of code at any given moment.

## 3.2) Profiling



The *Rprof* function will show you how much time is spent in
different functions, which can help you pinpoint bottlenecks in your
code. 

Here's a function that works with a correlation matrix such as one
might have for time series data. Basically, it creates a matrix of time lags (*dd*) and
computes the correlation between the outcome for all pairs of times, based on the time
lag between the pair of times. Then it computes the Cholesky factor of the correlation 
matrix so that it can generate a random time series in the last line. The question
one might ask is which part(s) of the code take the most time.

```{r, Rprof, eval=FALSE}
makeTS <- function(param, len){
	times <- seq(0, 1, length = len)
	dd <- rdist(times)
	C <- exp(-dd/param)
	U <- chol(C)
	white <- rnorm(len)
	return(crossprod(U, white))
}

library(fields)
Rprof("makeTS.prof")
out <- makeTS(0.1, 1000)
Rprof(NULL)
summaryRprof("makeTS.prof")
```

Here's the result for the *makeTS* function from the demo code
file:

```
$by.self   self.time self.pct total.time total.pct 
".Call"         0.38    48.72       0.38     48.72 
".Fortran"      0.22    28.21       0.22     28.21 
"matrix"        0.08    10.26       0.30     38.46 
"exp"           0.08    10.26       0.08     10.26 
"/"             0.02     2.56       0.02      2.56
$by.total         total.time total.pct self.time self.pct 
"makeTS"                0.78    100.00      0.00     0.00 
".Call"                 0.38     48.72      0.38    48.72 
"chol.default"          0.38     48.72      0.00     0.00 
"chol"                  0.38     48.72      0.00     0.00 
"standardGeneric"       0.38     48.72      0.00     0.00 
"matrix"                0.30     38.46      0.08    10.26 
"rdist"                 0.30     38.46      0.00     0.00 
".Fortran"              0.22     28.21      0.22    28.21 
"exp"                   0.08     10.26      0.08    10.26 
"/"                     0.02      2.56      0.02     2.56
$sample.interval [1] 0.02
$sampling.time [1] 0.78
```

*Rprof* tells how much time was spent in each function alone
(the *self* columns); this does not include time spent in
function calls made within the function. It also shows the aggregate time spent in
a function and all of the functions that it calls (the *total*
columns). Usually the former is going to be more useful, but in some
cases we need to decipher what is going on based on the latter.
The two different tables (*$by.self* and *$by.total*) show essentially the same information but 
sorted either based on the self percentages or the total percentages.

The *time* columns show clock time in seconds, while the *pct* columns
show the percentage of the time spent. 

Let's figure out what is going on here. In the first table, the self time tells us that
*.Call* (a call to C code) and *.Fortran* (a call to Fortran
code) take up most of the time, followed by *matrix* and *exp*. Looking at the
total time and seeing in *chol.default* that *.Call* is
used (you would have to go in and look at the *La_chol* function
to verify that *.Internal* uses *.Call*) and in *rdist* that *.Fortran* and
*matrix* are used we can infer that about 49% of the time
is being spent in the Cholesky and 38% in the *rdist* calculation,
with 10% in *exp*. As we increase the number of time points,
the time taken up by the Cholesky would increase since that calculation
is order of $n^{3}$ while the others are order $n^{2}$.

In this case, since the Cholesky and the main calculations in *rdist*, as well as *exp*,
are all done in compiled C or Fortran code, there is probably not much we can do
to speed this up (apart from using an optimized BLAS, which is essential). But in other cases profiling may reveal the slow steps in a piece of code. 

Note that *Rprof* works by sampling - every little while during a calculation it finds out what function R is in and saves that information to the file given as the argument to *Rprof*. So if you try to profile code that finishes really quickly, there's not enough opportunity for the sampling to represent the calculation accurately and you may get spurious results.

You might also check out some packages that help with profiling: *proftools*,
*profr*, and *lineprof*.

*Warning*: *Rprof* conflicts with threaded linear algebra,
so you may need to set OMP_NUM_THREADS to 1 to disable threaded
linear algebra if you profile code that involves linear algebra. 


# 4) Strategies for improving efficiency


## 4.1) Pre-allocate memory

It is very inefficient to iteratively add elements to a vector, matrix,
data frame, array or list (e.g., using *c*, *cbind*,
*rbind*, etc. to add elements one at a time). Instead, create the full object in advance
(this is equivalent to variable initialization in compiled languages)
and then fill in the appropriate elements. The reason is that when
R appends to an existing object, it creates a new copy and as the
object gets big, most of the computation involves the repeated 
memory allocation to create the new objects.  Here's
an illustrative example, but of course we would not fill a vector
like this using loops because we would in practice use vectorized calculations.

```{r, preallocate}
```

It's not necessary to use *as.numeric* above though it saves
a bit of time. **Challenge**: figure out why I have `as.numeric(NA)`
and not just `NA`.

In some cases, we can speed up the initialization by initializing a vector of length one and then changing its length and/or dimension, although in  many practical
circumstances this would be overkill.

For example, for matrices, start with a vector of length one, change the length, and then change the
dimensions

```{r, init-matrix}
```

For lists, we can do this

```{r, init-list}
```



## 4.2) Vectorized calculations

One key way to write efficient R code is to take advantage of R's
vectorized operations.

```{r, vectorize, cache=TRUE}
```

So what is different in how R handles the calculations above that
explains the huge disparity in efficiency? The vectorized calculation is being done natively
in C in a for loop. The explicit R for loop involves executing the for
loop in R with repeated calls to C code at each iteration. This involves a lot
of overhead because of the repeated processing of the R code inside the loop.

You can
usually get a sense for how quickly an R call will pass things along
to C or Fortran by looking at the body of the relevant function(s) being called
and looking for *.Primitive*, *.Internal*, *.C*, *.Call*,
or *.Fortran*. Let's take a look at the code for `+`,
*mean.default*, and *chol.default*. 

```{r, primitive}
```

Many R functions allow you to pass in vectors, and operate on those
vectors in vectorized fashion. So before writing a for loop, look
at the help information on the relevant function(s) to see if they
operate in a vectorized fashion. Functions might take vectors for one or more of their arguments.

```{r, vectorized}
```

**Challenge**: Consider the chi-squared statistic involved in
a test of independence in a contingency table:
\[
\chi^{2}=\sum_{i}\sum_{j}\frac{(y_{ij}-e_{ij})^{2}}{e_{ij}},\,\,\,\, e_{ij}=\frac{y_{i\cdot}y_{\cdot j}}{y_{\cdot\cdot}}
\]
where $y_{i\cdot}=\sum_{j}y_{ij}$ and $y_{\cdot j} = \sum_{i} y_{ij}$. Write this in a vectorized way
without any loops.  Note that 'vectorized' calculations also work
with matrices and arrays.

Vectorized operations can sometimes be faster than built-in functions 
(note here the *ifelse* is notoriously slow),
and clever vectorized calculations even better, though sometimes the
code is uglier. Here's an example of setting all negative values in a 
vector to zero.

```{r, vec-tricks, cache=TRUE}
```


Additional tips:
 - If you do need to loop over dimensions of a matrix or array, if possible
loop over the smallest dimension and use the vectorized calculation
on the larger dimension(s). For example if you have a 10000 by 10 matrix, try to set
up your problem so you can loop over the 10 columns rather than the 10000 rows.
 - In general, looping over columns is likely to be faster than looping over rows
given R's column-major ordering (matrices are stored in memory as a long array in which values in a column are adjacent to each other).
 - You can use direct arithmetic operations to add/subtract/multiply/divide
a vector by each column of a matrix, e.g. `A*b` does element-wise multiplication of
each column of *A* by a vector *b*. If you need to operate
by row, you can do it by transposing the matrix. 

Caution: relying on R's recycling rule in the context of vectorized
operations, such as is done when direct-multiplying a matrix by a
vector to scale the rows relative to each other, can be dangerous as the code is not transparent
and poses greater dangers of bugs. In some cases you may want to
first write the code transparently and
then compare the more efficient code to make sure the results are the same. It's also a good idea to  comment your code in such cases.

## 4.3) Using *apply* and specialized functions

Another core efficiency strategy is to use the *apply* functionality.
Even better than *apply* for calculating sums or means of columns
or rows (it also can be used for arrays) is {row,col}{Sums,Means}.

```{r, apply}
```

We can 'sweep' out a summary statistic, such as subtracting
off a mean from each column, using *sweep*

```{r, sweep}
```

Here's a trick for doing the sweep based on vectorized calculations, remembering
that if we subtract a vector from a matrix, it subtracts each element
of the vector from all the elements in the corresponding ROW. Hence the 
need to transpose twice. 

```{r, vectorized-sweep}
```

### Are *apply*, *lapply*, *sapply*, etc. faster than loops?

Using *apply* with matrices and versions of *apply* with lists may or may not be faster
than looping but generally produces cleaner code. Whether looping
is slower will depend on whether a substantial part of the work is
in the overhead of the looping or in the time required by the function
evaluation on each of the elements. If you're worried about speed,
it's a good idea to benchmark the *apply* variant against looping.

Here's an example where *apply* is not faster than a loop. Similar
examples can be constructed where *lapply* or *sapply* are not faster
than writing a loop. 

```{r, apply-vs-for}
```


## 4.4) Matrix algebra efficiency

Often calculations that are not explicitly linear algebra calculations
can be done as matrix algebra. For example, we can sum the rows of a matrix by multiplying by a vector of ones. It's actually even faster with
*rowSums*, so it's not a great example, but this sort of trick
does come in handy in surprising places. 

```{r, matrix-calc}
```

On the other hand, big matrix operations can be slow. **Challenge**: Suppose you
want a new matrix that computes the differences between successive
columns of a matrix of arbitrary size. How would you do this as matrix
algebra operations? It's possible to write it as multiplying the matrix
by another matrix that contains 0s, 1s, and -1s in appropriate places.
 Here it turns out that the
*for* loop is much faster than matrix multiplication. However,
there is a way to do it faster as matrix direct subtraction. 

When doing matrix algebra, the order in which you do operations can
be critical for efficiency. How should I order the following calculation?

```{r, linalg-order, cache=TRUE}
```

Why is the second order much faster?

We can use the matrix direct product (i.e., `A*B`) to do
some manipulations much more quickly than using matrix multiplication.
**Challenge**: How can I use the direct product to find the trace
of a matrix, $XY$? 

Finally, when working with diagonal matrices, you can generally get much faster results by being smart. The following operations: $X+D$, $DX$, $XD$
are mathematically the sum of two matrices and products of two matrices.
But we can do the computation without using two full matrices.
**Challenge**: How?

```{r, diag}
```

More generally, sparse matrices and structured matrices (such as block
diagonal matrices) can generally be worked with MUCH more efficiently
than treating them as arbitrary matrices. The R packages *spam* (for arbitrary
sparse matrices), *bdsmatrix* (for block-diagonal matrices),
and *Matrix* (for a variety of sparse matrix types) can help, as can specialized code available in other languages,
such as C and Fortran packages.


## 4.5) Fast mapping/lookup tables

Sometimes you need to map between two vectors. E.g., 
$y_{ij}\sim\mathcal{N}(\mu_{j},\sigma^{2})$
is a basic ANOVA type structure, where multiple observations in group $j$ 
are associated with a common mean, $\mu_j$. 

How can we quickly look up the mean associated with each observation?
A good strategy is to create a vector, *grp*, that gives a numeric
mapping of the observations to their cluster. Then you can access
the $\mu$ value relevant for each observation as: `mus[grp]`. This requires
that *grp* correctly map to the right elements of *mus*.

The *match* function can help in creating numeric indices that can then be used for lookups. 
Here's how you would create an index vector, *grp*, if it doesn't already exist.

```{r, match-lookup}
```

R allows you to look up elements of vector by name.
For example:

```{r, name-lookup}
```

You can do similar things in terms of looking up by name with dimension
names of matrices/arrays, row and column names of dataframes, and
named lists.

However, looking things up by name can be slow relative to looking up by index.
Here's a toy example where we have a vector or list with a million elements and
the character names of the elements are just the character versions of the 
indices of the elements.  

```{r, index-lookup}
```

Lookup by name is slow because R needs to scan through the objects
one by one until it finds the one with the name it is looking for.
In contrast, to look up by index, R can just go directly to the position of interest.


In contrast, we can look up by name in an environment very quickly, because environments in R use hashing, which allows for fast lookup that does not require scanning through all of the names in the environment. In fact, this is how R itself looks for values when you specify variables in R code. 

```{r, env-lookup, cache=TRUE}
```

The first benchmark indicates that lookup in an environment is nearly as fast as lookup by index in a vector or list, though the microbenchmark suggests that lookup in the environment is somewhat slower. 


